|

### 2020-12-29 paper/v1_old/broadband.speed.covid.Rmd
abstract + directories

---
title: "Working from home and digital divides: resilience in the time of a pandemic"
author:
- name: A. Uthor*
  num: 1,2
- name: O. Tro
  num: 2
- name: O. Vriga
  num: 3
address:
- num: 1
  org: Department of Incredible Research, University A, City A, Country A
- num: 2
  org: Department of Applied Things, University B, City B, Country B
- num: 3
  org: Very Important Stuff Committee, Institute C, City C, Country C
corrauth: "Corresponding author name, This is sample corresponding address."
email: correspondingauthor@protonmail.com
abstract: "TBC"
keywords:   covid; internet; working from home; broadband speed; time-series clusters;
classoption:
  - Royal
  - times
bibliography: bibliography
bibliographystyle: sageh
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{dcolumn}
  \def\tightlist{}
  \newcommand{\beginappendix}{\setcounter{table}{0} \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
output:
  rticles::sage_article:
    keep_tex: yes
---

@@

During the pandemic, working from home using digital technologies, whether partially
or exclusively, was transformed from a niche means of accessing work, albeit one that
had been on a slow, upward trend, to a widespread way of life in many countries.
The ability to work from home or telecommute meant millions retained their jobs and,
to a varying extent, maintained productivity during periods of strict
lockdowns around the world.
However, this ability has not been evenly distributed socially or spatially,
creating new intersections of economic and digital divisions, never tested before.
On one side are those who can work from home, supported by digital technologies,
and have thus been able to enjoy both economic resilience and greater personal safety.
On the other side, previously employed individuals have been forced to accept
furlough or redundancy packages unless they are part of the cadre of essential
workers, who are potentially at high risk of infection.
Whilst the basis for this pandemic-generated divide has been viewed as mainly 
occupational, here we consider whether it is also technological.

<!-- our approach -->
Using the UK as a case study, this paper aims to understand how the quality and
reliability of internet service, as reflected in *experienced* internet speeds,
may reinforce or redress the spatial and social dimensions of digital divisions.
To do so, we employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK.
Our analysis enables us to better understand how the spatial and social distribution
of both occupations and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand.
We will also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era,
and broadband services and infrastructure must be fit for purpose.

@@

Yet the shift towards telecommuting during various stages of lockdown around the world
has been drastic and there are speculations that post-Covid, the tendency to work from
home will be much higher than pre-Covid, raising questions around whether internet
services can accommodate the increased demand.
For example, $47$% of people in employment in the UK worked solely from home in April
$2020$, whilst the same figure only reached $5$% the year before [@ons2020;
@ons2020lm2019].
A back of the envelope calculation suggests that up to 40% of the labour force
could work from home on an ongoing basis [@batty2020editorial].
Similar figures have been reported for other countries [@felstead2020homeworking].
Approximately $37$% of the European workforce worked from home in April $2020$ with
countries like Finland reaching $60$% [@eurofound2020].
In the US, almost half of the working population worked from home during the same
period because of the pandemic [@brynjolfsson2020covid],
and a recent estimate indicated that $37$% of all jobs in the US can be permanently
performed entirely from home [@NBERw26948].
None of these changes could have happened in the absence of reliable information
and communication technology (ICT) infrastructure -- both in terms of software
and hardware.
But while software innovations are easily diffused across space and
society^[See for example the huge success of videoconferencing apps such as Zoom
[@marks2020zoom].], the same does not apply for ICT hardware infrastructure such as 
internet broadband connectivity.

<!-- ICT infrastructure and digital divides-->
The literature describes first level digital divides in terms of the
availability and quality of internet connectivity, such as that manifest in
different geographies in the UK [@riddlesden2014broadband; @philip2017digital].
Second level digital divides consider the presence or lack of the necessary skills
to effectively utilise digital technologies and the internet
[@blank2014dimensions; @van2011internet].
The third level focuses on the heterogenous returns of internet usage among
different socioeconomic groups and, consequently, how digital technologies can
assist in bridging or further enhancing existing socioeconomic divides.
[@stern2009levels; @van2014digital; @van2015third].
The capability to telecommute is related to all three levels of digital divides,
but more importantly leads to differentiated outcomes regarding the economic
resilience of people and places to overcome a systemic shock such as the current
pandemic.
The extreme level of demand for telecommuting fundamentally alters the
potential returns of internet use for the user and wider community,
assumes skills or functions that are present in some occupations but not in others,
and relies upon access to high quality and reliable internet services.

As the quality of internet infrastructure and services, as well as the concentration
of different occupations are spatially dependent and clustered in space, 
our approach offers a framework for understanding the impact of and interactions 
between digital divisions geographically and socioeconomically.
By asking how resilient broadband speeds, and particularly upload speeds, are
experienced in different parts of the UK during a time of extreme demand,
we interrogate which places benefit from the greater economic resilience
digital technologies can offer, not only during the pandemic,
but also into the future.
The structure of this paper is as follows.
First we review the literature on telecommuting and digital divides to better
understand their structural and spatial development pre-pandemic,
and thus their importance to the economic resilience of different places.
We then describe our data and methodology.
Our results section first offers a classification of how internet services vary
across clusters of UK local authorities and then assesses whether these clusters 
replicate or repudiate other socio-economic and geographic patterns of economic 
resilience.
We conclude with a discussion of the insights we've gained from our new perspective 
on digital divisions.

@@

<!-- Who works from home-->
In this analysis, the terms ‘telecommuting’ and ‘working from home^
[See also the popular social media hashtag #WFM]’ are used
interchangeably, as most remote labour during the Covid-19 crisis was carried out
in the homes of individual employees rather than any other location [@eurofound2020].
However, it should be noted that previous research has explored how telecommuting
can occur in other places, including satellite offices or on public transport
[@felstead2012rapid; @siha2006telecommuting].
Previous research has also used a variety of definitions to measure the level of
telecommuting within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or part-time, and
those who use digital technologies to work remotely full-days or part-days
[@allen2015effective; @bailey2002review; @haddad2009examination].
No matter the definition, the option and capability to telecommute or work from home
has never been equally distributed spatially or socio-economically any more than
different industries and employment opportunities have.
For example, studies from the United States, the Netherlands, and the UK found
that telecommuters are most likely to hold professional, managerial, and technical
occupations where the workforce is better educated and wealthier, and that there is
suppressed demand among women and part-time workers
[@headicar2016move; @peters2004employees; @singh2013modeling].

@@

## Digital divides and economic resilience {#sec:2.2}
<!-- WFH and digital divides-->
Our understanding of telecommuting as a product of enabled occupations can be
described as a manifestation of the third level digital divide, as those who are able
to use digital technologies to work from home benefit from a high rate of return on
their use of the internet in terms of autonomy, flexibility, and time saved from
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
These returns have been even greater during the Covid-19 crisis,
when those with the capability to telecommute also have the ability to maintain
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first
level digital divide, which is associated with access to and quality of internet
connectivity.
@SALEMINK2017360 provides a systematic review of the pre-pandemic,
first level digital divide in infrastructure quality between urban and rural areas
in various advanced economies.
Rural areas, predictably, fare worse.
Yet whether this variation in infrastructure quality affects the spatial footprint of
telecommuting has not previously been measured, in part because telecommuting has
not previously been the cause of greatest demand and pressure on internet services.

There are indications that those who purchase high speed connections consume
more data of all sorts and use their connections for a variety of purposes
[@hauge2011consumer].
There is also a correlation between access to internet services and a
reduction in household transport spend [@bris2017ict].
Independently of whether the additional internet use and reduced travel is
because of increased telecommuting, these studies suggest that better internet
services enable households to make savings and efficiencies,
an example of the first level digital divide reinforcing the third level.
Yet did the purchase of high speed connections and increased internet access also
prepare households for long-term home-working, enforced by government restrictions?
The extreme demand during the pandemic provides a new opportunity to understand
how infrastructure accessibility, quality, and reliability affects telecommuting,
particularly in light of the high volumes of bandwidth-intensive video conferencing
required in order to avoid the face-to-face contact that could increase
the spread of infection.
We seek to answer how internet service resilience contributes to or reduces economic 
resilience when the latter is dependent upon the capability to work from home.
We also aim to improve our understanding of the impact of first level digital 
division on telecommuting, and whether this results in much more fundamental third 
level digital division than has previously been perceived.

<!-- Covid-19 and working from home, cities, urban structure -->
Furthermore, these multi-layered digital divides intersect with material divides
and the economic geography of the UK.
Following the regional economic resilience literature, which underlines the 
differentiated capacity of cities and regions to escape or recover from economic 
crises [@martin2012regional; @kitsos2018economic], different places have different 
industrial and occupational profiles, and these affect the aggregated potential 
capacity of places for telecommuting.
Such profiles are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide [@martin_north_south].
Various studies have illustrated severe inequalities between the north and the
south regions of England regarding, for example, skills and human capital,
unemployment, productivity and prosperity [@lee2014grim; @mccann2020perceptions;
@dorling2018peak].
Some scholars have even argued that the UK suffers some of the highest level of
interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
Not only are all three levels of digital divides associated with or shaped by the 
geography of the UK, but the intersection of the digital and material divides 
affects the capacity of places to overcome some of the economic effects of the 
Covid-19 pandemic.
Importantly, this is the first time that digital technologies became an essential
tool for economic resilience for such a great part of the population.

@@

## Time-Series Clustering {#sec:3.1}
<!-- clustering in general -->
First we identified our methodological framework as cluster analysis, which can
be defined within the modern machine learning framework as an unsupervised
learning task, partitioning unlabelled observations into homogeneous groups
known as clusters [@montero2014tsclust].
The key idea is that observations within clusters tend to be more similar than
observations between clusters.
Clustering is particularly useful for exploratory studies as it identifies
structures within the data [@aghabozorgi2015time].
Therefore, cluster analysis is a widely used family of techniques in geography
[@gordon1977classification; @everitt1974cluster].
For instance, clustering methods are the basis of *geodemographics*, a research 
domain which aims to create small area indicators or typologies of neighbourhoods
based on diverse variables [@SINGLETON2009289; @harris2005geodemographics].
Clustering techniques have also been employed to solve *regionalisation* problems
[@niesterowicz2016].
Common characteristics of these studies are the cross-sectional nature of the data
they employ.
Indeed, most clustering problems in geography deal with observations that are
fixed in time.
However, for this paper we are interested in internet speeds, which vary over time.
We therefore create clusters of local authorities in the UK with similar temporal
signatures of experienced internet speeds, and employ time-series clustering methods.

<!-- time-series clustering -->
Time-series clustering methods have been developed in order to deal with clustering
problems linked to, for instance, stock or other financial data, economic,
governmental or medical data as well as machine monitoring
[@aggarwal2013time; @aggarwal2001surprising; @hyndman2015large; @WARRENLIAO20051857].
The main challenge -- and also the difference with cross-sectional clustering
problems -- is data dimensionality given the multiplicity of data points for every
individual object, local authorities in our case, included in the data set.
Time-series are dynamic data as the value of the observations change as
a function of time [@aghabozorgi2015time].
This high dimensionality leads to
(i) computational and algorithmic challenges regarding handling these data and
building algorithms to perform clustering over long time-series, and
(ii) open questions regarding the choice of similarity measures in order to cluster
similar times series objects together considering the whole length of the time-series
and overcoming issues around noise, outliers and shifts [@lin2004iterative;
@aghabozorgi2015time].

For this paper we utilise a category of time-series clustering methods
known as shape-based approaches.
These methods match two sperate time-series objects based on the similarity of
their shapes through the calculation of distances between the shapes,
and are better equipped to capture similarities between short length time-series 
[@aghabozorgi2015time], such as our data.
We thus identify clusters of UK local authorities with similar temporal signatures --
i.e. shapes -- of experienced internet speeds.
The clusters are identified using a partitioning algorithm, which leads to 
non-overlapping clusters, and particularly the widely use *k*-means algorithm because
of the simplicity of the implementation and the interpretability of the
results.
This is an iterative algorithm, which begins with defining the desired number of
clusters *k*.
Then each observation is randomly assigned to a cluster from the $[1,k]$ space.
This initial cluster assignment is followed by iterations in order to minimise
the distance between the centroids of the clusters and the observations
assigned to these clusters [@james2013introduction].

<!-- difference between clustering cross-sectional and time-series data -->
There are a number of differences between the application of *k*-means for 
cross-sectional data and its application for times series data.
Instead of creating clusters around centroids, a common approach is to create
clusters around *medoids*, which are representative time-series objects with a
minimal distance to all other cluster objects [@sardatime].
Also, instead of calculating the Euclidean distance between centroids and data
points, more complex distance measures need to be employed in order to capture
the similarity between a time-series object and a medoid.
A common distance measure for shape-based time-series clustering is
Dynamic Time Warping (DTW), an algorithm comparing two time-series objects to find 
the optimum warping path between them.
DTW is widely used in order to overcome limitations linked to the use of Euclidean
distance [@sardatime; @berndt1994using; @ratanamahatana2004everything].
The `R` package `dtwclust` has been used for the time-series clustering [@dtwclust].

## Experienced Broadband Speeds {#sec:3.2}
<!-- introduction of our broadband speeds data -->
To assess the internet quality and reliability across local authorities in the UK, 
we utilise unique data comprising individual internet speed tests from
Speedchecker Ltd^[[https://www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)].
This is a private company that allows internet users to check their own broadband
upload and download speeds, and stores every speed-check with timestamp and
geolocation information.
These data have been used before to assess digital divides [@riddlesden2014broadband]
and the impact of local loop unbundling regulatory processes 
[@nardotto2015unbundling], and we followed the former's approach to remove outliers. 
By using this volunteered geographic data, we are able to assess the internet speed 
*experienced* by users, which may differ from the maximum speeds *advertised* by 
ISPs.
We are particularly interested in upload speeds and the frequency of speed tests
over the period from March to May $2020$, as government statements indicate this is
when UK workers were first told to work from home if at all possible [@GovUK2020].
Average upload speeds are slower than average download speeds, at $9.3$Mb/s
mean upload speed for the whole sample during the period of interest,
compared to $29.6$Mb/s for download speeds, but they are also less associated
with internet-based, high-demand, leisure activities such as video streaming.
Therefore, upload speeds are more relevant to work-related activities such as
uploading documents or two-way audio, video, and text-based communication systems.

<!-- data transformation -->
The frequency of speed tests was important in identifying the temporal profile which 
would give us most insight into experienced  internet service and resilience over 
units of time.
Our analysis indicates that whilst there is an overall trend of increased testing 
from March to April and then a slight reduction from April to May,
this trend masks substantial variation by not only the day of the week,
but also time of day, as can be seen in Figure \ref{test2020}.
Thus, a daily aggregation of upload speeds would mask the variation in experienced
service over the course of each weekday.
Furthermore, the importance of this variation is highlighted by a comparison
with the same period in $2019$, as in Figure \ref{test2019}, when the volume of 
testing and thus of experience of internet service quality peaked in the evening, 
presumably in response to demand for leisure activities and download speeds.
In contrast, the majority of the increase in testing in $2020$ is during the working
day, creating a new morning peak in Figure \ref{test2020}.
Therefore, we include a measure of hourly variation in our temporal profiles to
reflect the change in users' perception of the workday reliability of internet 
services.

However, we judged that there were insufficient observations -- only $631$ speed 
tests per LAD on average -- for each for each working hour of each working day in 
each Local Authority District (LAD) to profile speeds at that level of detail.
Spatial aggregation was also necessary because we could not follow individuals or
households and connect data points.
Therefore, we aggregate the $241,088$ individual, geolocated and time-stamped 
speed-checks during the $13$ weeks of March to May inclusive for weekdays in $2020$ 
by each hour of the day and day of the week.
As our research aims to identify the geography of internet service resilience for
work purposes, bank holidays and the hours between midnight and $6:00$ were excluded,
as well as weekend days.
The composite week time-series thus comprise $18$ hours multiplied by $5$ weekdays or
$90$ time points per series.
The time-series were calculated for each of the $382$ LADs in the UK, standardised,
and then a *k*-means partitioning around medoids clustering algorithm
was applied using DTW.
We initially run the algorithm for $k \in \mathbb{N} \bigcap [5,15]$
and used cluster validity indices (CVIs) to pick the optimal solution of $k = 13$.
Following @sardatime the majority vote for the following CVIs was used: Silhouette
(max), Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min), Modified
Davies-Bouldin (DB*, min), Dunn (max), COP (min).

@@

\pagebreak

# Appendix 1 {#appendix1}

This is the LAD cluster membership for the upload speed timeseries.

```{r echo=FALSE, message=FALSE, results = "asis"}

# Cluster table
clusters <- paste(path, "/data/temp/clusters_nodiff.csv", sep = "")
clusters <- read_csv(clusters)
clusters <- clusters %>%
  dplyr::select(LAD, cluster.up) %>%
  #rename('cluster membership' = cluster.up) %>%
  #group_by('cluster membership') %>%
  group_split(cluster.up, .keep = F)
  #dplyr::arrange()
  #sort('cluster membership', decreasing = F)

>>>>

---
title: |
  WFH and broadband speed (title needs rework)
type: ARTICLE TEMPLATE
author:
  - name: A. N. Author
    affil: a
    email: latex.helpdesk@tandf.co.uk
  - name: John Smith
    affil: b
    email: john.smith@uni-freiburg.de
affiliation:
  - num: a
    address: |
      Taylor & Francis, 4 Park Square, Milton Park, Abingdon, UK
  - num: b
    address: |
      Institut für Informatik, Albert-Ludwigs-Universität, Freiburg, Germany
bibliography: bibliography.bib
appendix: appendix.tex
abstract: |
  TBC
keywords: |
  covid; internet; working from home; broadband speed; time-series clusters
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{dcolumn}
  \def\tightlist{}
output: rticles::tf_article
---

@@

During the pandemic, working from home using digital technologies, whether partially
or exclusively, was transformed from a niche means of accessing work, albeit one that
had been on a slow, upward trend, to a widespread way of life in many countries.
The ability to work from home or telecommute meant millions retained their jobs and,
to a varying extent, maintained productivity during periods of strict
lockdowns around the world.
However, this ability has not been evenly distributed socially or spatially,
creating a new type of digital divide.
On one side are those who can work from home, supported by digital technologies,
and have thus been able to enjoy both economic resilience and greater personal safety.
On the other side, previously employed individuals have been forced to accept
furlough or redundancy packages unless they are part of the cadre of essential
workers, who are potentially at high risk of infection.
Whilst the basis for this new digital divide has been viewed as mainly occupational,
here we consider whether the divide is also technological.

<!-- our approach -->
Using the UK as a case study, this paper aims to understand how the quality and
reliability of internet service, as reflected in *experienced* internet speeds,
may reinforce or redress the spatial and social dimensions of the digital division
exposed by the pandemic.
To do so, we employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK.
Our analysis enables us to better understand how the spatial and social distribution
of both occupations and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand.
We will also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era,
and broadband services and infrastructure must be fit for purpose.
**LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ**

@@

Yet the shift towards telecommuting during various stages of lockdown around the world
has been drastic and there are speculations that post-Covid, the tendency to work from
home will be much higher than pre-Covid, raising questions around whether internet
services can accommodate the increased demand.
For example, $47$% of people in employment in the UK worked solely from home in April
$2020$, whilst the same figure only reached $5$% the year before [@ons2020;
@ons2020lm2019].
A back of the envelope calculation suggests that up to 40% of the working force
could work from home on an ongoing basis [@batty2020editorial].
Similar figures have been reported for other countries [@felstead2020homeworking].
For instance, $37$% of the European workforce worked from home in April $2020$ with
countries like Finland reaching $60$% [@eurofound2020].
In the US, almost half of the working population worked from home during the same
period because of the pandemic [@brynjolfsson2020covid],
and a recent estimate indicated that $37$% of all jobs in the US can be permanently
performed entirely from home [@NBERw26948].

<!-- ICT infrastructure and digital divides-->
None of these changes could have happened in the absence of reliable information
and communication technology (ICT) infrastructure -- both in terms of software
and hardware.
But while software innovations are easily diffused across space and
society^[See for example the huge success of videoconferencing apps such as Zoom
[@marks2020zoom].],
the same does not apply for ICT hardware infrastructure such as internet
broadband connectivity.
The literature describes first level digital divides in terms of the
availability and quality of internet connectivity, such as that manifest in
different geographies in the UK [@riddlesden2014broadband; @philip2017digital].
Second level digital divides consider the presence or lack of the necessary skills
to effectively utilise digital technologies and the internet
[@blank2014dimensions; @van2011internet].
The third level focuses on the heterogenous returns of internet usage among
different socioeconomic groups and, consequently, how digital technologies can
assist in bridging or further enhancing existing socioeconomic divides.
[@stern2009levels; @van2014digital; @van2015third].
The capability to telecommute is related to all three levels of digital divides,
but more importantly leads to differentiated outcomes regarding the economic
resilience of people and places to overcome a systemic shock such as the current
pandemic.
The extreme level of demand for telecommuting fundamentally alters the
potential returns of internet use for the user and wider community,
assumes skills or functions that are present in some occupations but not in others,
and relies upon access to high quality internet services.

As the quality of internet infrastructure and services, as well as variation in
occupations are spatially dependent and clustered in space, our approach offers a
framework for understanding the impact of and interactions between
the different levels of digital division in different places
with different characteristics.
By asking how resilient broadband speeds, and particularly upload speeds are
as experienced in different parts of the UK during a time of extreme demand,
we interrogate which places benefit from the greater economic resilience
digital technologies can offer, not only during the pandemic,
but also into the future.
The structure of this paper is as follows.
First we review the literature on telecommuting and digital divides to better
understand the structural and spatial development of these practices pre-pandemic,
and thus their importance to the economic resilience of different places.
We then describe our data and methodology.
Our results section first offers a classification of how internet services vary
across the UK local authorities and then assesses whether these clusters replicate
or repudiate other socio-economic and geographic patterns of economic resilience.
The paper ends with a conclusions section.

@@

<!-- Who works from home-->
In this analysis, the terms ‘telecommuting’ and ‘working from home’ are used
interchangeably, as most remote labour during the Covid-19 crisis was carried out
in the homes of individual employees rather than any other location [@eurofound2020].
However, it should be noted that previous research has explored how telecommuting
can occur in other places, including satellite offices or on public transport
[@felstead2012rapid; @siha2006telecommuting].
Previous research has also used a variety of definitions to measure the level of
telecommuting within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or part-time, and
those who use digital technologies to work remotely full-days or part-days
[@allen2015effective; @bailey2002review; @haddad2009examination].
No matter the definition, the option and capability to telecommute or work from home
has never been equally distributed spatially or socio-economically any more than
different industries and employment opportunities have.
For example, studies from the United States, the Netherlands, and the UK found
that telecommuters are most likely to hold professional, managerial, and technical
occupations where the workforce is better educated and wealthier, and that there is
suppressed demand among women and part-time workers
[@headicar2016move; @peters2004employees; @singh2013modeling].

@@

## Digital divides and economic resilience {#sec:2.2}
<!-- WFH and digital divides-->
Our understanding of telecommuting as a product of enabled occupations can be
described as a manifestation of the third level digital divide, as those who are able
to use digital technologies to work from home benefit from a high rate of return on
their use of the internet in terms of autonomy, flexibility, and time saved from
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
These returns have been even greater during the Covid-19 crisis,
when those with the ability to telecommute also have the ability to maintain
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first
level digital divide, which is associated with access and quality of internet
connectivity at a time of extremely high demand.
@SALEMINK2017360 provides a systematic review of the pre-pandemic,
first level digital divide in infrastructure quality between urban and rural areas
in various advanced economies.
Rural areas, predictably, fare worse.
Yet whether this variation in infrastructure quality affects the spatial footprint of
telecommuting has not previously been measured, in part because telecommuting has
not previously been the cause of greatest demand and pressure on internet services.

There are indications that those who purchase high speed connections consume
more data of all sorts and use their connections for a variety of purposes
[@hauge2011consumer],
and that there is a correlation between access to internet services and a
reduction in household transport spend [@bris2017ict].
Whether the implication is more internet use and less travel
because of increased telecommuting, these studies suggest that better internet
services enable households to make savings and efficiencies,
an example of the first level digital divide reinforcing the third level.
Yet did the purchase of high speed connections and increased internet access also
prepare households for long-term home-working, enforced by government restrictions?
The extreme demand during the pandemic provides a new opportunity to understand
how infrastructure accessibility, quality, and reliability affects telecommuting,
particularly in light of the high volumes of bandwidth-intensive video conferencing
required in order to avoid the face-to-face contact that could increase
the spread of infection.
We seek to answer how internet service resilience contributes to or reduces
economic resilience when the latter is dependent upon
the capability to work from home.
We also aim to improve our understanding of the impact
of first level digital division on telecommuting, and whether this results in
much more fundamental third level digital division than has previously been perceived.

<!-- Covid-19 and working from home, cities, urban structure -->
Furthermore, these multi-layered digital divides intersect with material divides
and the economic geography of the UK.
Following the regional economic resilience literature, which
underlines the differentiated capacity of cities and regions to escape or
recover from economic crises [@martin2012regional; @kitsos2018economic],
different places have different industrial and occupational profiles,
and these affect the aggregated potential capacity of places for telecommuting.
Such profiles are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide [@martin_north_south].
Various studies have illustrated severe inequalities between the north and the
south regions of England **replace 'UK'** regarding, for example, skills and human capital,
unemployment, productivity and prosperity [@lee2014grim; @mccann2020perceptions;
@dorling2018peak].
Some scholars have even argued that the UK suffers some of the highest level of
interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
Not only are all three levels of digital divides associated, to a certain extent,
with or shaped by the geography of the UK, but the intersection of the digital and
material divides affects the capacity of places to overcome
some of the economic effects of the Covid-19 pandemic.
Importantly, this is the first time that digital technologies became an essential
tool for economic resilience for such a great part of the population.

@@

## Time-Series Clustering {#sec:3.1}
<!-- clustering in general -->
The starting point of our methodological framework is cluster analysis, which can
be defined within the modern machine learning framework as an unsupervised
learning task, partitioning unlabelled observations into homogeneous groups
known as clusters [@montero2014tsclust].
The key idea is that observations within clusters tend to be more similar than
observations between clusters.
Clustering is particularly useful for exploratory studies as it identifies
structures within the data [@aghabozorgi2015time].
Therefore, cluster analysis is a widely used family of techniques in geography
[@gordon1977classification; @everitt1974cluster].
For instance, clustering methods are the basis of *geodemographics*, a research domain
which aims to create small area indicators or typologies of neighbourhoods
based on various and sometimes diverse variables
[@SINGLETON2009289; @harris2005geodemographics].
Clustering techniques have also been employed to solve *regionalisation* problems
[@niesterowicz2016].

<!-- time-series clustering -->
Common characteristics of these studies are the cross-sectional nature of the data
they employ.
Indeed, most clustering problems in geography deal with observations that are
fixed in time.
However, for this paper we are interested in internet speeds, which vary over time
and, therefore, create clusters of local authorities in the UK with similar temporal
signatures of experienced internet speeds.
Hence, we deviate from the established geographical clustering tools and
employ time-series clustering methods.

Time-series clustering methods have been developed in order to deal with clustering
problems linked to, for instance, stock or other financial data, economic,
governmental or medical data as well as machine monitoring
[@aggarwal2013time; @aggarwal2001surprising; @hyndman2015large; @WARRENLIAO20051857].
The main challenge -- and also the difference with cross-sectional clustering
problems -- is data dimensionality given the multiplicity of data points for every
individual object, local authorities in our case, included in the data set.
Time-series are dynamic data as the value of the observations change as
a function of time [@aghabozorgi2015time].
This high dimensionality leads to
(i) computational and algorithmic challenges regarding handling these data and
building algorithms to perform clustering over long time-series, and
(ii) open questions regarding the choice of similarity measures in order to cluster
similar times series objects together considering the whole length of the time-series
and overcoming issues around noise, outliers and shifts [@lin2004iterative;
@aghabozorgi2015time].

<!-- Time-series clustering methods utilising the whole length of time-series can be grouped in three categories. The first -- model-based approaches -- is based on recovering the underlying model for each time-series and then applying clustering algorithms on the model parameters of each time-series [@aghabozorgi2015time]. The main criticism is the cluster accuracy for nearby clusters [@mitsa2010temporal].The second approach is based the formation of vectors of features derived from the original time-series. These new data of reduced dimensionality are then clustered using conventional clustering algorithms. -->

For this paper we utilise a category of time-series clustering methods
known as shape-based approaches.
These methods match two sperate time-series objects based on the similarity of
their shapes through the calculation of distances between the shapes,
and are thus better equipped to capture similarities between short length
time-series [@aghabozorgi2015time].
This approach serves best this paper because (i) we identify clusters of
UK local authorities with similar temporal signatures -- i.e. shapes -- of
experienced internet speeds and (ii) the length of our time-series is short
(see the data discussion in this section).

Another import element of time-series clustering is the actual clustering algorithm.
Similar to the clustering of cross-sectional data, we can employ partitioning
algorithms, which lead to non-overlapping clusters, hierarchical clustering,
which classifies clusters at different levels, and fuzzy algorithms, which
create overlapping clusters [@sardatime].
Because of the simplicity of the implementation and the interpretability of the
results, we utilise here partitioning clustering based on the widely used *k*-means
algorithm.
This is an iterative algorithm, which begins with defining the desired number of
clusters *k*.
Then each observation is randomly assigned to a cluster from the $[1,k]$ space.
This initial cluster assignment is followed by iterations in order to minimise
the distance between the centroids of the clusters and the observations
assigned to these clusters [@james2013introduction].

<!-- difference between clustering cross-sectional and time-series data -->
There are a number of differences between the above described application of
*k*-means for cross-sectional data and its application for times series data.
Instead of creating clusters around centroids, a common approach is to create
clusters around *medoids*, which are representative time-series objects with a
minimal distance to all other cluster objects [@sardatime].
Also, instead of calculating the Euclidean distance between centroids and data
points, more complex distance measures need to be employed in order to capture
the similarity between a time-series object and a medoid.
A common distance measure for shape-based time-series clustering is
Dynamic Time Warping (DTW).
Using its underpinning dynamic programming algorithm, DTW compares two time-series
objects to find the optimum warping path between them.
DTW is widely used in order to overcome limitations linked to the use of Euclidean
distance [@sardatime; @berndt1994using; @ratanamahatana2004everything].
The `R` package `dtwclust` has been used for the time-series clustering [@dtwclust].

## Experienced Broadband Speeds {#sec:3.2}
<!-- introduction of our broadband speeds data -->
To assess the quality and reliability of internet across local authorities
in the UK during the time when the population were told to work from home if at
all possible we utilise unique data comprising individual internet speed tests from
Speedchecker Ltd^[[https://www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)].
This is a private company that allows internet users to check their own broadband
upload and download speeds, and stores every speed-check with timestamp and
geolocation information.
These data have been used before to assess digital divides [@riddlesden2014broadband]
and the impact of local loop unbundling regulatory processes [@nardotto2015unbundling].
By using volunteered geographic data, we are able to assess the *experienced*
internet speed by users, which may differ from the *advertised* maximum speeds of ISPs.

We are particularly interested in upload speeds and the frequency of speed tests
over the period from March to May $2020$, as government statements indicate this is
when UK workers were told to work from home if at all possible [@GovUK2020].
Average upload speeds are slower than average download speeds, at $9.3$Mb/s
mean upload speed for the whole sample during the period of interest,
compared to $29.6$Mb/s for download speeds, but they are also less associated
with internet-based, high-demand, leisure activities such as video streaming.
Therefore, upload speeds are more relevant to work-related activities such as
uploading documents or two-way audio, video, and text-based communication systems.
Meanwhile, the frequency of speed tests was important in identifying
the temporal profile which would give us most insight into
experienced internet service and resilience, and provide
an indication of the volume of experience over particular units of time.

<!-- data transformation -->
The first step in the workflow after dropping some outliers following
@riddlesden2014broadband was to transform the individual, geolocated and time-stamped
tests to more meaningful aggregates both in terms of space and time.
The frequency of testing indicates that whilst there is an overall trend of increased
testing from March to April and then a slight reduction from April to May,
this trend masks substantial variation by not only the day of the week,
but also time of day, as can be seen in Figure \ref{test2020}.
Thus, a daily aggregation of upload speeds would mask the variation in experienced
service over the course of each weekday.
Furthermore, the importance of this variation is highlighted by a comparison
with the same period in $2019$, as in Figure \ref{test2019},
when the volume of testing and thus of experience of internet service quality
was much more tied to the evening leisure peak and presumably to download speeds.
Since the increase in testing during the working day in $2020$ is an indication that
users have a greater perception of the variable quality of internet service,
particularly during a new morning peak of testing for service reliability,
we decided to include a measure of hourly variation in our temporal profiles.

However, although this is a large data set -- $241,088$ individual tests performed
during weekday hours across the study period --
there are not enough observations for each Local Authority District (LAD)
and for each working hour of each working day -- $631$ speed tests per LAD on average
-- to profile speeds at that level of detail.
Therefore, we aggregate all the speed-checks during the $13$ weeks of March to May
inclusive for weekdays in $2020$ by each hour of the day and day of the week.
As our research aims to identify the geography of internet service resilience for
work purposes, bank holidays and the hours between midnight and $6:00$ were excluded,
as well as weekend days.
The composite week time-series thus comprise $18$ hours multiplied by $5$ weekdays or
$90$ time points per series.
We also aggregate these data spatially because we could not follow individuals or
households and connect data points.
The time-series were calculated for each of the $382$ LADs in the UK, standardised,
and then a *k*-means partitioning around medoids clustering algorithm
was applied using DTW.
We initially run the algorithm for $k \in \mathbb{N} \bigcap [5,15]$
and used cluster validity indices (CVIs) to pick the optimal solution of $k = 13$.
Following @sardatime the majority vote for the following CVIs was used: Silhouette
(max), Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min), Modified
Davies-Bouldin (DB*, min), Dunn (max), COP (min).

@@
